---
title: "NHANES project challenge"
author: "Eliana Ibrahimi, Alise Ponsero, Marta B. Lopes"
date: "2023-10-13"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The problem

Sleep disorder is a highly prevalent disorder among US adults and
despite advancements in treatment options, prevalence rates are
increasing. We want to assess the relationship between sleep disorder
and predictors which are reported to impact sleep disorder. 

## The dataset

A sample of n = 24118 adults from the National Health and Nutrition
Examination Survey (NHANES) from 2008 to 2018.

#### Variables and codding

Sleep disorder: 1=Yes, 0=No

BMI numeric

Age numeric

BMI categorical: 2=Normal, 1=Underweight, 3=Overweight, 4=Obese

Age categorical: 1=18-29, 2=30-49, 3=50-69, 4= ≥70

Gender: 1=Male, 2=Female

Marital status: 1=Married/Living with Partner, 2=Widowed,
3=Divorced/Separated, 4=Never married

Race: 1=Hispanic, 2=White, 3=Black, 4=Other

Diet: 1=Good to Excellent, 2=Poor to fair,

Physical Activity (Vigorous), 1=Yes, 2=No

Smoking: 1=Current smoker, 2=Former smoker, 3=Never smoker

Mental health (depression): 1=Not at all, 2=Several days, 3=More than
half the days, 4=Nearly every day

General health: 1=Very good to excellent, 2=Good, 3=Fair to poor

```{r}
#Load the data

data=read.csv("data_nhanes_msb.csv")
data
```

### The challenge

### 
