---
title: "Data Exploration"
author: "Eliana Ibrahimi"
date: "2023-10-04"
output: html_document
---

## Introduction to R

"R is an environment, based on S plus language, within which many classical and modern statistical techniques have been implemented. A few of these are built into the base R environment, but many are supplied as packages. There are about 25 packages supplied with R and many more are available through the CRAN family of Internet sites (via https://CRAN.R-project.org) and elsewhere. Most classical statistics and much of the latest methodology is available for use with R, but users may need to be prepared to do a little work to find it." (W. N. Venables, D. M. Smith and the R Core Team., 2018) - Check the R-intro document in Open Source Materials/Hands-on Biostatistics in Drive for more infortmation. 


## Start with R

### Working directory

1. Create a sub-directory named "R" in your "Documents" folder. 

2. From RStudio, use the menu to change your working directory under Session > Set Working Directory > Choose Directory. Choose the directory you've just created in step 1


You can run the code below to set working directory also.

```{r}
#setwd("C:/Users/user/Desktop/HOB(2020)/R") #set working directory

#getwd() # Shows the working directory (wd)
#setwd(choose.dir()) # Select the working directory interactively
#setwd("C:/myfolder/data") # Changes the wd
#setwd("H:\\myfolder\\data") # Changes the wd
```




### Installing/loading packages/

R packages are collections of functions and data sets developed by the community. 

```{r, include=FALSE}

#Install the following packages:

#install.packages("foreign")
#library(foreign) #load the package 
install.packages("car")
#install.packages("Hmisc")
#install.packages("reshape")

#Install multiple packages at once

#install.packages(c("slidify", "ggplot2", "devtools"))


#load package

library(car) #will load the car package

#http://cran.r-project.org/web/views/
```


### Getting help

"Before asking others for help, it's generally a good idea for you to try to help yourself. R includes extensive facilities for accessing documentation and searching for help. There are also specialized search engines for accessing information about R on the internet, and general internet search engines can also prove useful" (https://www.r-project.org/help.html). 

```{r}
?plot # Get help for an object. You can also type: help(plot)

#??regression # Search the help pages for anything that has the word "regression". You can also type:

#help.search("regression")


#help(package=car) # View documentation in package car. You can also type: library(help="car")

```


## Read/import data in R 

### Option 1

You can add your data as vectors and create the dataframe using the data.frame() function as follows.

```{r}


# Data in two numeric vectors

women_weight <- c(38.9, 61.2, 73.3, 21.8, 63.4, 64.6, 48.4, 48.8, 48.5)
men_weight <- c(67.8, 60, 63.4, 76, 89.4, 73.3, 67.3, 61.3, 62.4) 

# Create a data frame

my_data <- data.frame( 
  group = rep(c("Woman", "Man"), each = 9), 
  weight = c(women_weight,  men_weight)
)

# Saving all objects to file *.RData
save.image("my_data.RData") 

```



### Option 2: Import or read your data from excel, csv.

```{r}

# Read csv file above my_data
library(readr)
my_data <- read_csv("my_data.csv")
View(my_data)


# Read csv file from PC storage 
#my_data2<- read.csv("C:/Users/user/Desktop/R/my_data.csv")

#Read excel file
#library(readxl) #Load the readxl package, if not installed, do that before.
#my_data3<- read_excel("C:/Users/user/Desktop/HOB(2020)/R/my_data.xlsx")

#Data can be imported from Environment-Import dataset as well

```


### Option 3: Load data from R

```{r}

# Load the data saved as RData in option 1
load("my_data.RData") #Add path to data if necessary


```


## Export/write the data in other formats

```{r}

#Write the dataframe as csv file
write.csv(my_data, "C:/Users/user/Desktop/HOB(2020)/R/my_data.csv")


#Write the dataframe as .txt file
write.table(my_data, file = "my_data.txt", sep = "\t")

```

## Data exploration

```{r}

summary(my_data) # Provides basic descriptive statistics and frequencies.
edit(my_data) # Open data editor
str(my_data) # Provides the structure of the dataset
names(my_data) # Lists variables in the dataset
head(my_data) # First 6 rows of dataset
head(my_data, n=4)# First 4 rows of dataset
head(my_data, n= -3) # All rows but the last 3
tail(my_data) # Last 6 rows
tail(my_data, n=5) # Last 5 rows
tail(my_data, n= -5) # All rows but the first 5
my_data[1:5, ] # First 5 rows
my_data[1:5,1:2] # First 5 rows of data of the first 2 variables

```



### Missing data

```{r}
rowSums(is.na(my_data)) # Number of missing per row
colSums(is.na(my_data)) # Number of missing per column/variable

# Convert to missing data
my_data[my_data$weight=="& ","weight"] <- NA # NOTE: Notice hidden spaces.
my_data[my_data$weight==999,"weight"] <- NA

# The function complete.cases() returns a logical vector indicating which cases are complete.
# list rows of data that have missing values
my_data[!complete.cases(my_data),]

# The function na.omit() returns the object with listwise deletion of missing values.
# Creating a new dataset without missing data
my_data1 <- na.omit(my_data)
```


### Value labels/recode variables

```{r}

#Read data from ecxel. 
library(readxl)
my_data_num <- read_excel("my_data.num.xlsx")


# Use factor() for nominal data
my_data_num$group <- factor(my_data_num$group, levels = c(1,2), labels = c("male", "female"))


# Use ordered() not factor() for ordinal data
```



### Creating ids/sequence of numbers

```{r}
# Creating a variable with a sequence of numbers from 1 to n (where ‘n’ is the total number of observations)
my_data$id <- seq(dim(my_data)[1])
my_data
```


### Recoding variables/creating categories

```{r}
library(car)
my_data$weight.rec <- recode(my_data$weight, 
"30:50='30-50';
51:70='51-70';
71:90='71-90'")
my_data$weight.rec <- as.factor(my_data$weight.rec)
```

### Sort data; Deleting variables

```{r}

#Sort data by weight.rec
my_data.sorted <- my_data[order(my_data$weight.rec),]

#Delete variables
my_data$weight.rec <- NULL


```


### Subseting the data

```{r}
mydata3 <- subset(my_data, weight >= 20 & weight <= 50)
mydata4 <- subset(my_data, weight >= 20 & weight <= 50, select=c(id, weight))

mydata5 <- subset(my_data, group=="Woman" & weight >= 70)
mydata6 <- subset(my_data, group=="Woman" & weight == 70)
```


### Categorical data: Frequencies/Crosstabs

```{r}

#Frequencies
table(my_data$group)


# Two-way tables/ crosstabs

library(readr)
data <- read_csv("chi-square.csv", 
    col_types = cols(id = col_number(), improvement = col_character(), 
        treatment = col_character()))

#Create crosstab
dt<- table(data$treatment, data$improvement)
dt

addmargins(dt) # Adding row/col margins

#Calculate proportions
round(prop.table(dt,1), 2) # Round col prop to 2 digits
round(100*prop.table(dt,1), 2) # Round col prop to 2 digits (percents)
addmargins(round(prop.table(dt,1), 2),2) # Round col prop to 2 digits

round(prop.table(dt,2), 2) # Round column prop to 2 digits
round(100*prop.table(dt,2), 2) # Round column prop to 2 digits (percents)
addmargins(round(100*prop.table(dt,2), 2),1) # Round col prop to 2 digits

round(prop.table(dt),2) # Tot proportions rounded
round(100*prop.table(dt),2) # Tot proportions rounded
```

### Numerical data

Descriptive Statistics

```{r}

summary(my_data) # Summary of all numeric variables

mean(my_data$weight) #mean
median(my_data$weight) #median
var(my_data$weight) # Variance
sd(my_data$weight) # Standard deviation
max(my_data$weight) # Max value
min(my_data$weight) # Min value
range(my_data$weight) # Range
quantile(my_data$weight) # Quantiles 25%
quantile(my_data$weight, c(.3,.6,.9)) # Customized quantiles


length(my_data$weight) # Num of observations when a variable is specify
length(my_data$weight) # Number of variables when a dataset is specify

table(my_data$group)
names(sort(-table(my_data$group)))[1]

```


Descriptive statistics by groups

```{r}
# Descriptive statistics by groups using --tapply--
mean <- tapply(my_data$weight,my_data$group, mean, na.rm=TRUE)
sd <- tapply(my_data$weight,my_data$group, sd)
median <- tapply(my_data$weight,my_data$group, median)
max <- tapply(my_data$weight,my_data$group, max)

table <- round(cbind(mean, median, sd, max),digits=1)
table
```



## Confidence intervals for the mean


```{r}

#install.packages(distributions3)

library(distributions3) # load package

# The data
x <- c(8.5, 9.3, 7.9, 9.2, 10.3)
n<-length(x)

  
# t-student with 4 degrees of freedom
T_4 <- StudentsT(df = 4)

# 95% CI
L1= mean(x) - quantile(T_4, 1-0.05 / 2) * sd(x) / sqrt(n)
L2= mean(x) + quantile(T_4, 1 - 0.05 / 2) * sd(x) / sqrt(n)
# 

L1 #7.917
L2 # 10.163

```




## Graphs

### Histograms

```{r}
x <- my_data$weight

hist(x, freq=F, col="red", breaks = 5)

help("hist")

```


### Grouped histograms
```{r}

hist(my_data$weight, breaks="FD", col="green", main="Histogram of weight")
hist(my_data$weight[my_data$group=="Man"], breaks="fd", col="gray", add=TRUE)
legend("topright", c("Woman","Man"), fill=c("green","gray"))


```

## Other graphs

See other lectures where the data is visualized with appropriate graphs before each statistical model. 



## Exploring the workspace

```{r}
objects() # Lists the objects in the workspace
ls() # Same as objects()
remove() # Remove objects from the workspace
rm(list=ls()) #clearing memory space
search() # Shows the loaded packages
library() # Shows the installed packages
dir() # show files in the working directory
```


## References

(https://www.rdocumentation.org/packages)

http://www.sthda.com/english/wiki/running-rstudio-and-setting-up-your-working-directory-easy-r-programming

http://www.stat.auckland.ac.nz/~paul/RGraphics/rgraphics.html

http://addictedtor.free.fr/graphiques/

http://addictedtor.free.fr/graphiques/thumbs.php?sort=votes

http://www.statmethods.net/advgraphs/layout.html

http://socserv.mcmaster.ca/jfox/

Quick R http://www.statmethods.net/

UCLA Resources to learn and use R http://www.ats.ucla.edu/stat/R/

https://www.r-bloggers.com

https://statistics.laerd.com/
