---
title: "Multiple linear regression"
author: "Eliana ibrahimi"
date: "2023-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset

Data from the [Maryland Biological Stream Survey](http://dnr.maryland.gov/streams/Pages/mbss.aspx)

The dependent variable is the number of longnose dace (*Rhinichthys cataractae*) per 75-meter section of stream (fish).

The independent variables are the area (in acres) drained by the stream; the dissolved oxygen (in mg/liter); the maximum depth (in cm) of the 75-meter segment of the stream; nitrate concentration (mg/liter); sulfate concentration (mg/liter); and the water temperature on the sampling date (in degrees C), and hardness (\<45=0 mg equivalent CACO3/L and \>45=1), as dummy variable. A dummy variable is a numerical variable that is used in a regression analysis to "code" for a binary categorical variable.

Biological goal might be to measure the physical and chemical characteristics of a stream and be able to predict the abundance of longnose dace; another goal might be to generate hypotheses about the causes of variation in longnose dace abundance.

### Read the data in R

```{r}
library(readr)
fish_data <- read_csv("fish_data.csv")
fish_data$hardness=as.factor(fish_data$hardness)

```

#### Explore the data

##### Linear relationship

```{r}
 ## Scatterplot
pairs(fish_data, pch = 18, col = "steelblue")

## Boxplot of response/dependent variable

boxplot(fish_data$fish)

#We could remove the extreme outliers

 #fish_data2=subset(fish_data, fish<160)

 boxplot(fish_data2$fish)
```

```{r}

#install.packages(plot3D)
library(plot3D)
attach(fish_data)

y=fish
x=acre
z=no3

# Create a scatter plot
 scatter3D(x, y, z, phi = 0, bty = "g",
        pch = 20, cex = 2, ticktype = "detailed")
# Add another point (black color)
scatter3D(x = 7, y = 3, z = 3.5, add = TRUE, colkey = FALSE, 
         pch = 18, cex = 3, col = "black")


# type ="h" for vertical lines
scatter3D(x, y, z, phi = 0, bty = "g",  type = "h", 
           ticktype = "detailed", pch = 19, cex = 0.5)


```

## Perform multiple linear regression

### Building the model

```{r }
#Building the model 

#attach(fish_data)

lm<-lm(fish ~ acre + do2 +	depth +	no3 +	so4 +	temp +	hardness, data = fish_data)

summary(lm)

confint(lm) #extract CI of the coefficients
```

##Model selection based on AIC

```{r}
##Model selection based on AIC
library(MASS)
lm2 <- stepAIC(lm)
lm2$anova
```

### Fit the final model

```{r}
#Fit the final model
lm_final<-lm(fish ~ acre + depth + no3 + hardness, data=fish_data)
summary(lm_final)
```

## Check Model assumptions

#### Normality of residuals

```{r}
# Get the model residuals
model_residuals = lm_final$residuals

# Plot the result
hist(model_residuals)
```

The histogram looks skewed to the left; hence we can not conclude the normality with enough confidence. Instead of the histogram, let's look at the residuals along the normal Q-Q plot. If there is normality, then the values should follow a straight line.

```{r}
# Plot the residuals
qqnorm(model_residuals)
# Plot the Q-Q line
qqline(model_residuals)
```

From the plot, we can observe that a few portions of the residuals do not lie in a straight line. Then we can not surely assume that the residuals of the model do not follow a normal distribution.

```{r}
shapiro.test(model_residuals)
#not normally distribted
```

#### Multicollinearity assumption check

```{r}
# Install and load the ggcorrplot package
#install.packages("ggcorrplot")
library(ggcorrplot)

# Remove the Customer Value column
reduced_data <- subset(fish_data, select = -fish)
# To avoid an error we need to have hardness as numeric
reduced_data$hardness=as.numeric(reduced_data$hardness)

# Compute correlation at 2 decimal places
corr_matrix = round(cor(reduced_data), 2)

# Compute and show the  result
ggcorrplot(corr_matrix, hc.order = TRUE, type = "lower",
          lab = TRUE)
```

No strong correlation among predictors (i.e., \>\|0.7\|)

##### **Another way to do this...**

```{r}
#install and load the GGally library
#install.packages("GGally")
library(GGally)

#generate the pairs plot
ggpairs(reduced_data)
```

### Check for homoskedasticity

The variance of the residuals should be consistent for all observations. This preferred condition is known as homoskedasticity. Violation of this assumption is known as [heteroskedasticity](https://www.statology.org/understanding-heteroscedasticity-in-regression-analysis/).

To check if this assumption is met we can create a *fitted value vs. residual plot:*

```{r}
#create fitted value vs residual plot
plot(fitted(lm_final), residuals(lm_final))

#add horizontal line at 0
abline(h = 0, lty = 2)
```

Ideally we would like the residuals to be equally scattered at every fitted value. We can see from the plot that the scatter tends to become a bit larger for larger fitted values, but this pattern isn't extreme enough to cause too much concern.

## Interpreting the results of the model

```{r}
#Fit again the final model
lm_final<-lm(fish ~ acre + depth + no3 + hardness, data=fish_data)
summary(lm_final)
```

-   Model: fish abundance = -3.225 + 0.00177(acre) + 0.384(depth) + 7.164 (no3) - 0.357 (hardness_high)

-   The overall F-statistic of the model is 11.42 and the corresponding p-value is **4.931e-07**. This indicates that the overall model is statistically significant. In other words, the regression model as a whole is useful.

-   acre is statistically significant at the 0.01 significance level. In particular, the coefficient from the model output tells is that a one unit increase in acre is associated with a 0.00017 unit increase, on average, in number of fish, assuming other predictors are held constant.

-   depth is statistically significant at the 0.05 significance level. In particular, the coefficient from the model output tells is that a one unit increase in depth is associated with a 0.384 unit increase, on average, in fish abundance, assuming other predictors are held constant.

-   Interpret yourself no3 ....

-   hardness -3.576e+01 is the difference in fish abundance between a high level of hardness compared to the low level. The negative value indicates that the fish abundance is lower in high levels of hardness than in low levels of hardness.

```{r}
plot(fish_data$fish~fish_data$hardness)
```

### **Assessing the Goodness of Fit of the Model**

To assess how "good" the regression model fits the data, we can look at a couple different metrics:

**1. Multiple R-Squared**

This measures the strength of the linear relationship between the predictor variables and the response variable. A multiple R-squared of 1 indicates a perfect linear relationship while a multiple R-squared of 0 indicates no linear relationship whatsoever.

In this example, the multiple R-squared is 0.42. Adjusted R-squared is 0.3836. This indicates that **38.36%** of the variance in *fish abundance* can be explained by the predictors in the model.

**Related:** [What is a Good R-squared Value?](https://www.statology.org/what-is-a-good-r-squared-value/)

### **Using the Model to Make Predictions**

```{r}
#define new observation
#0.00177(acre) + 0.384(depth) + 7.164 (no3) - 0.357 (hardness_high)
newdata = data.frame(acre=5000, depth=150, no3=1.5, hardness="1")

#use model to predict points value
predict(lm_final, newdata)

```

Predicted fish abuncance is 38.24

## Plot the predictions against the observed values

```{r}
predictions= predict(lm_final)

plot(fish~predictions)
```
